---
title: "PCN-ARRS diagnostic snapshot"
format:
  typst:
    papersize: a4
    margin:
      x: 1.8cm
      y: 1.6cm
params:
  pcn_code: "U24992" # Heritage PCN — flagged as a staffing outlier (outside 95% band) by the national funnel-plot check
---

```{r}
#| label: setup
#| include: false

library(targets)
library(dplyr)
library(ggplot2)
library(tibble)

source_dir <- "../src"
for (f in list.files(source_dir, full.names = TRUE)) source(f)

# Reads pre-built pipeline outputs — run `targets::tar_make()` from the
# project root before rendering this report. Nothing is recomputed here.
store <- "../_targets"
tr <- function(name) tar_read_raw(name, store = store)

national_pcn_list_size <- tr("national_pcn_list_size")
national_pcn_primary_icb <- tr("national_pcn_primary_icb")
national_pcn_workforce <- tr("national_pcn_workforce")
national_staffing_funnel_flags <- tr("national_staffing_funnel_flags")

pcn_code <- params$pcn_code

pcn_name <- filter(national_pcn_list_size, PCN_CODE == pcn_code)$PCN_NAME
icb_name <- filter(national_pcn_primary_icb, PCN_CODE == pcn_code)$ICB_NAME

list_size_row <- filter(national_pcn_list_size, PCN_CODE == pcn_code)
workforce_row <- filter(national_pcn_workforce, PCN_CODE == pcn_code)
funnel_row <- filter(national_staffing_funnel_flags, PCN_CODE == pcn_code)

# The headline claim a PCN is identified because its staffing is a
# statistical outlier, not because it happens to rank highest — so the
# copy has to read differently depending on whether this PCN actually
# is one (direction points from lower-than-expected FTE for the ARRS
# case; a small minority of PCNs land here — see
# national_staffing_funnel_flags for the national split).
flag_sentence <- dplyr::case_when(
  funnel_row$funnel_flag == "outside_99.8pct" &
    funnel_row$direction == "understaffed_vs_expected" ~
    "That gap is large enough to sit outside the 99.8% range — the strongest, most statistically defensible case for the next ARRS hire.",
  funnel_row$funnel_flag == "outside_95pct" &
    funnel_row$direction == "understaffed_vs_expected" ~
    "That gap sits outside the 95% range where staffing this far below expectation is unlikely to be chance — worth a serious look at the next ARRS hire.",
  funnel_row$funnel_flag == "within_expected_range" ~
    "That's within the range of normal variation nationally — not a statistical outlier on its own.",
  TRUE ~
    "That's actually above what's expected given its patient population — the opposite direction, worth understanding before assuming a staffing gap here."
)

# Palette — dataviz skill's validated reference instance
# (references/palette.md). Swap for a brand palette by changing these
# constants only.
col_accent <- "#2a78d6" # categorical slot 1 (blue)
col_muted <- "#c3c2b7" # baseline/axis gray, used for de-emphasised bars
col_ink_primary <- "#0b0b0b"
col_ink_secondary <- "#52514e"
col_ink_muted <- "#898781"
col_gridline <- "#e1e0d9"
col_good <- "#0ca30c"
col_critical <- "#d03b3b"

chart_theme <- theme_minimal(base_size = 9) +
  theme(
    text = element_text(colour = col_ink_secondary),
    axis.text = element_text(colour = col_ink_muted, size = 7.5),
    axis.title = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = col_gridline, linewidth = 0.3),
    plot.title = element_text(colour = col_ink_primary, size = 10, face = "bold"),
    plot.subtitle = element_text(colour = col_ink_secondary, size = 8),
    legend.position = "none"
  )

pct1 <- function(x) sprintf("%.1f%%", x)
```

# `{r} pcn_name`

`{r} icb_name` &nbsp;·&nbsp; `{r} format(list_size_row$list_size, big.mark = ",")` registered patients &nbsp;·&nbsp; rendered `{r} format(Sys.Date())`

```{r}
#| label: headline
#| echo: false
#| output: asis

cat("::: {.callout-important appearance=\"simple\"}\n\n")
cat(sprintf(
  "**%s is staffed at %.0f%% of what its age profile and deprivation predict.** %.1f FTE in post against an expected %.1f FTE, benchmarked against every PCN in England (Spiegelhalter funnel plot — see below). %s\n\n",
  pcn_name, funnel_row$staffing_ratio_pct, funnel_row$actual_fte, funnel_row$expected_fte,
  flag_sentence
))
cat(":::\n")
```

## At a glance

| Metric | This PCN | What it means |
|---|---|---|
| Staffing vs. expected need | `{r} sprintf("%.0f%%", funnel_row$staffing_ratio_pct)` (`{r} gsub("_", " ", funnel_row$funnel_flag)`) | Actual FTE against what the PCN's age profile and deprivation predict, benchmarked nationally — a PCN is flagged for being a statistical outlier, not for topping a ranking |
| ARRS FTE currently in post | `{r} sprintf("%.1f", workforce_row$arrs_fte)` | Additional-roles staff already funded and working in this PCN |

## Is this staffing gap real, or just noise?

```{r}
#| label: fig-staffing-funnel
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: !expr paste0("Actual vs. expected FTE across all ", nrow(national_staffing_funnel_flags), " PCNs nationally (Spiegelhalter overdispersion-adjusted control limits). ", pcn_name, " marked in blue.")

funnel_data <- national_staffing_funnel_flags |>
  mutate(is_this_pcn = PCN_CODE == pcn_code)

this_row <- filter(funnel_data, is_this_pcn)

# Flip label alignment depending on which half of the (log) x-range the
# highlighted point falls in, so the direct label never runs off the
# panel edge regardless of which PCN this report is parameterised for.
x_mid_log <- mean(log10(range(funnel_data$expected_fte)))
label_hjust <- if (log10(this_row$expected_fte) > x_mid_log) 1.1 else -0.1

ggplot(funnel_data, aes(x = expected_fte, y = staffing_ratio_pct)) +
  geom_hline(yintercept = this_row$od95_lower_pct, colour = col_ink_muted, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od95_upper_pct, colour = col_ink_muted, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od99_8_lower_pct, colour = col_ink_muted, linetype = "dotted", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od99_8_upper_pct, colour = col_ink_muted, linetype = "dotted", linewidth = 0.3) +
  geom_point(aes(colour = is_this_pcn, size = is_this_pcn), alpha = 0.7) +
  geom_text(
    data = this_row, aes(label = pcn_name),
    hjust = label_hjust, size = 2.8, colour = col_ink_primary, fontface = "bold"
  ) +
  scale_colour_manual(values = c(`TRUE` = col_accent, `FALSE` = col_muted)) +
  scale_size_manual(values = c(`TRUE` = 2.6, `FALSE` = 1)) +
  scale_x_log10(
    labels = scales::label_number(scale_cut = scales::cut_short_scale()),
    expand = expansion(mult = c(0.12, 0.12))
  ) +
  chart_theme +
  theme(axis.text = element_text(colour = col_ink_muted, size = 7.5))
```

Dashed lines mark the 95% control band, dotted lines the 99.8% band — both
computed nationally, so a PCN isn't flagged just because it's small and its
staffing ratio is noisy. "Expected" FTE reflects two things about this PCN's
own registered population: its age profile (patients under 15, 15–64, and
65+ each carry a different national-average staffing rate, since older
patients are well established to need more GP time) and its deprivation
profile — a standard indirect-standardisation approach (Spiegelhalter DJ,
"Funnel plots for comparing institutional performance," *Statistics in
Medicine* 2005;24(8):1185-1202). It doesn't need a peer group to make its
point: `{r} pcn_name` is plotted against every PCN in England, on the same
measure, with nothing else mixed in.

This model uses current (2026) national staffing and population data, and
its expected-staffing weights have been checked against peer-reviewed
academic research on GP consultation patterns (Mukhtar et al. 2018, *Br J
Gen Pract*, DOI 10.3399/bjgp18X695981).

# Limitations

- **This is a question, not a verdict.** This gap is worth a conversation,
  not a conclusion — case-mix, coding practice, and local context all vary
  in ways this data doesn't capture.
- **ARRS FTE gaps are treated as zero, not missing.** A handful of PCNs have
  no rows in NHS England Digital's Primary Care Network Workforce
  publication; this is treated as zero ARRS staff — the conservative read,
  since understating ARRS FTE only widens the staffing gap shown — rather
  than omitting the PCN, but the true figure could be higher than shown.
- **"Expected" FTE is a national average, not a funding entitlement.** It's
  built by fitting a Poisson rate regression of actual FTE against each
  PCN's own population (age profile, sex, and deprivation), with list size
  as the exposure — the same indirect-standardisation logic as NHS
  SMR-style funnel plots, but FTE is fractional, not a genuine count of
  discrete events, so the Poisson-based control limits are an
  approximation, not an exact statistical fit.
- **The age/deprivation weighting is our own fitted model, not NHS
  England's Global Sum Allocation Formula** (the "Carr-Hill formula," the
  standard NHS practice-funding formula). We could not obtain and verify
  its exact published age-sex coefficients, so instead we fit our own
  weights directly from national staffing data. Checked informally
  against Mukhtar TK, Bankhead C, Stevens S, Perera R, Holt TA, Salisbury
  C, Hobbs FDR, "Factors associated with consultation rates in general
  practice in England, 2013-2014," *Br J Gen Pract* 2018;68(670):e370-e377
  (a different outcome to what we model here, and not used as a
  production input) — deprivation terms corroborate closely, age terms
  diverge notably, an open question rather than a settled one. Doesn't
  include morbidity, which the literature treats as a further
  significant, independent driver of need that this pipeline doesn't yet
  capture.
- **No per-PCN £ figure for ARRS spend or underspend exists publicly**
  below ICB level — nothing here should be read as a costed claim.
- **DNA (Did Not Attend) rate and practice-level detail are tracked but
  not shown here** — this report is deliberately scoped to the ARRS
  staffing disclosure only; that detail is available in a follow-up
  conversation.
- **This report is a first draft**, rendered against one PCN as a test
  case. Page count, layout, and section order are all still open.
