---
title: "PCN-ARRS diagnostic snapshot"
format:
  typst:
    papersize: a4
    margin:
      x: 1.8cm
      y: 1.6cm
params:
  pcn_code: "U98580" # Hayes Wick PCN (SEL) — below the 99.8% band on ARRS-only staffing, June 2026 build
---

```{r}
#| label: setup
#| include: false

library(targets)
library(dplyr)
library(ggplot2)
library(tibble)

source_dir <- "../src"
for (f in list.files(source_dir, full.names = TRUE)) source(f)

# Reads pre-built pipeline outputs — run `targets::tar_make()` from the
# project root before rendering this report. Nothing is recomputed here.
store <- "../_targets"
tr <- function(name) tar_read_raw(name, store = store)

national_pcn_list_size <- tr("national_pcn_list_size")
national_pcn_primary_icb <- tr("national_pcn_primary_icb")
national_pcn_workforce <- tr("national_pcn_workforce")
national_staffing_funnel_flags <- tr("national_staffing_funnel_flags")
national_staffing_gap <- tr("national_staffing_gap")
national_arrs_role_benchmarks <- tr("national_arrs_role_benchmarks")
national_pcn_exclusions <- tr("national_pcn_exclusions")

pcn_code <- params$pcn_code

pcn_name <- filter(national_pcn_list_size, PCN_CODE == pcn_code)$PCN_NAME
icb_name <- filter(national_pcn_primary_icb, PCN_CODE == pcn_code)$ICB_NAME

list_size_row <- filter(national_pcn_list_size, PCN_CODE == pcn_code)
workforce_row <- filter(national_pcn_workforce, PCN_CODE == pcn_code)
funnel_row <- filter(national_staffing_funnel_flags, PCN_CODE == pcn_code)

# PCNs with no ARRS return are left out of the funnel entirely (see
# compute_funnel.R::pcn_staffing_data()) — there's no staffing comparison
# to report for them, so refuse to render rather than print an empty page.
if (nrow(funnel_row) == 0) {
  stop(pcn_code, " is not in the national staffing funnel (see ",
       "docs/pcn_exclusions.md) — no report to render.")
}

# Exclusion count for the Limitations section.
n_pcns_excluded <- nrow(national_pcn_exclusions)

# The headline claim a PCN is identified because its staffing is a
# statistical outlier, not because it happens to rank highest — so the
# copy has to read differently depending on whether this PCN actually
# is one (direction points from lower-than-expected FTE for the ARRS
# case; a small minority of PCNs land here — see
# national_staffing_funnel_flags for the national split).
flag_sentence <- dplyr::case_when(
  funnel_row$funnel_flag == "outside_99.8pct" &
    funnel_row$direction == "understaffed_vs_expected" ~
    "That gap is large enough to sit outside the 99.8% range — the strongest, most statistically defensible case for the next ARRS hire.",
  funnel_row$funnel_flag == "outside_95pct" &
    funnel_row$direction == "understaffed_vs_expected" ~
    "That gap sits outside the 95% range where staffing this far below expectation is unlikely to be chance — worth a serious look at the next ARRS hire.",
  funnel_row$funnel_flag == "within_expected_range" ~
    "That's within the range of normal variation nationally — not a statistical outlier on its own.",
  TRUE ~
    "That's actually above what's expected given its patient population — the opposite direction, worth understanding before assuming a staffing gap here."
)

# Palette — dataviz skill's validated reference instance
# (references/palette.md). Swap for a brand palette by changing these
# constants only.
col_accent <- "#2a78d6" # categorical slot 1 (blue)
col_muted <- "#c3c2b7" # baseline/axis gray, used for de-emphasised bars
col_ink_primary <- "#0b0b0b"
col_ink_secondary <- "#52514e"
col_ink_muted <- "#898781"
col_gridline <- "#e1e0d9"
col_good <- "#0ca30c"
col_critical <- "#d03b3b"

chart_theme <- theme_minimal(base_size = 9) +
  theme(
    text = element_text(colour = col_ink_secondary),
    axis.text = element_text(colour = col_ink_muted, size = 7.5),
    axis.title = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = col_gridline, linewidth = 0.3),
    plot.title = element_text(colour = col_ink_primary, size = 10, face = "bold"),
    plot.subtitle = element_text(colour = col_ink_secondary, size = 8),
    legend.position = "none"
  )

pct1 <- function(x) sprintf("%.1f%%", x)
```

# `{r} pcn_name`

`{r} icb_name` &nbsp;·&nbsp; `{r} format(list_size_row$list_size, big.mark = ",")` registered patients &nbsp;·&nbsp; rendered `{r} format(Sys.Date())`

```{r}
#| label: headline
#| echo: false
#| output: asis

cat("::: {.callout-important appearance=\"simple\"}\n\n")
cat(sprintf(
  "**%s has %.0f%% of the ARRS staff we would expect for its patients.** %.1f full-time-equivalent (FTE) ARRS staff in post, against an expected %.1f FTE, compared with every PCN in England (see below). %s\n\n",
  pcn_name, funnel_row$staffing_ratio_pct, funnel_row$actual_fte, funnel_row$expected_fte,
  flag_sentence
))
cat(":::\n")
```

## At a glance

| Metric | This PCN | What it means |
|---|---|---|
| ARRS staff vs. expected | `{r} sprintf("%.0f%%", funnel_row$staffing_ratio_pct)` (`{r} gsub("_", " ", funnel_row$funnel_flag)`) | ARRS staff in post, as a share of what PCNs serving similar patients typically have |
| ARRS FTE in post | `{r} sprintf("%.1f", workforce_row$arrs_fte)` | Staff the PCN employs in roles the Additional Roles Reimbursement Scheme (ARRS) can fund |

## Is this staffing gap real, or just noise?

We estimate how many ARRS staff a PCN would typically have, given how many
patients it serves, how old they are and how deprived the area is. We then
compare `{r} pcn_name`'s actual ARRS staff with that estimate, alongside
every other PCN in England. Small PCNs naturally vary more, so the normal
range (the funnel) is wider for them and narrows as PCNs get bigger. A PCN
outside it is unlikely to be there by chance.

```{r}
#| label: fig-staffing-funnel
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.1
#| fig-cap: !expr paste0("Each dot is one of ", nrow(national_staffing_funnel_flags), " PCNs in England. The solid line marks as expected, dashed lines the 95% normal range and dotted lines the 99.8% range. ", pcn_name, " is marked in blue.")

funnel_data <- national_staffing_funnel_flags |>
  mutate(is_this_pcn = PCN_CODE == pcn_code)

this_row <- filter(funnel_data, is_this_pcn)

# Flip label alignment depending on which half of the (log) x-range the
# highlighted point falls in, so the direct label never runs off the
# panel edge regardless of which PCN this report is parameterised for.
x_mid_log <- mean(log10(range(funnel_data$expected_fte)))
label_hjust <- if (log10(this_row$expected_fte) > x_mid_log) 1.1 else -0.1

ggplot(funnel_data, aes(x = expected_fte, y = staffing_ratio_pct)) +
  geom_hline(yintercept = 100, colour = col_ink_muted, linewidth = 0.3) +
  geom_hline(yintercept = this_row$od95_lower_pct, colour = col_ink_muted, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od95_upper_pct, colour = col_ink_muted, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od99_8_lower_pct, colour = col_ink_muted, linetype = "dotted", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od99_8_upper_pct, colour = col_ink_muted, linetype = "dotted", linewidth = 0.3) +
  geom_point(aes(colour = is_this_pcn, size = is_this_pcn), alpha = 0.7) +
  geom_text(
    data = this_row, aes(label = pcn_name),
    hjust = label_hjust, size = 2.8, colour = col_ink_primary, fontface = "bold"
  ) +
  scale_colour_manual(values = c(`TRUE` = col_accent, `FALSE` = col_muted)) +
  scale_size_manual(values = c(`TRUE` = 2.6, `FALSE` = 1)) +
  scale_x_log10(
    labels = scales::label_number(scale_cut = scales::cut_short_scale()),
    expand = expansion(mult = c(0.12, 0.12))
  ) +
  labs(
    x = "Expected ARRS FTE, given the PCN's patients (log scale)",
    y = "ARRS FTE in post, % of expected"
  ) +
  chart_theme +
  theme(axis.title = element_text(colour = col_ink_secondary, size = 7.5))
```

The normal range follows Spiegelhalter's standard funnel-plot method
(Spiegelhalter DJ, "Funnel plots for comparing institutional performance,"
*Statistics in Medicine* 2005;24(8):1185-1202). The expected figure comes
from a model fitted to June 2026 staffing and population data for every PCN
in England.

## How far is this from similar PCNs?

```{r}
#| label: fig-staffing-gap
#| echo: false
#| fig-width: 6.4
#| fig-height: 1.9
#| fig-cap: !expr paste0("How many more ARRS staff (FTE) ", pcn_name, " would have if it matched each national reference point. FTE is not a headcount, because many posts are part-time. Targets compare this PCN with PCNs that have similar patients; they are not a staffing requirement.")

gap_data <- filter(national_staffing_gap, PCN_CODE == pcn_code) |>
  mutate(
    benchmark = factor(benchmark, levels = rev(levels(benchmark))),
    gap_label = if_else(
      gap_fte > 0,
      sprintf("%.1f FTE  (+%.1f)", target_fte, gap_fte),
      sprintf("%.1f FTE  (already met)", target_fte)
    )
  )
actual_fte_now <- gap_data$actual_fte[1]
x_max <- max(gap_data$target_fte, actual_fte_now)

ggplot(gap_data, aes(y = benchmark)) +
  geom_vline(xintercept = actual_fte_now, colour = col_ink_muted, linewidth = 0.3) +
  geom_segment(
    aes(x = actual_fte, xend = target_fte, yend = benchmark),
    colour = col_accent, linewidth = 2, lineend = "round"
  ) +
  geom_point(aes(x = target_fte), colour = col_accent, size = 2.6) +
  geom_text(aes(x = target_fte, label = gap_label), hjust = -0.18, size = 2.6, colour = col_ink_primary) +
  annotate(
    "text", x = actual_fte_now, y = 4.55, label = sprintf("In post now: %.1f FTE", actual_fte_now),
    hjust = 1.05, size = 2.5, colour = col_ink_secondary
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.02))) +
  coord_cartesian(xlim = c(NA, x_max * 1.22), clip = "off") +
  labs(x = "ARRS FTE") +
  chart_theme +
  theme(
    panel.grid.major.y = element_blank(),
    axis.title.x = element_text(colour = col_ink_secondary, size = 7.5)
  )
```

## Which ARRS roles?

```{r}
#| label: fig-arrs-roles
#| echo: false
#| fig-width: 6.4
#| fig-height: 2.5
#| fig-cap: !expr paste0("ARRS staff (FTE) per 10,000 registered patients in each role, for ", pcn_name, " and for the typical (median) PCN in England. Roles that neither has are not shown. ARRS lets each PCN choose its own mix, so a lower figure in one role may reflect a choice, not a gap.")

role_data <- filter(national_arrs_role_benchmarks, PCN_CODE == pcn_code) |>
  filter(fte_per_10k > 0 | national_median_per_10k > 0) |>
  mutate(report_role = reorder(report_role, pmax(fte_per_10k, national_median_per_10k)))

series_levels <- c(pcn_name, "Typical PCN in England")
role_long <- bind_rows(
  transmute(role_data, report_role, value = fte_per_10k, series = series_levels[1]),
  transmute(role_data, report_role, value = national_median_per_10k, series = series_levels[2])
) |>
  mutate(series = factor(series, levels = series_levels))

ggplot(role_long, aes(x = value, y = report_role)) +
  geom_line(aes(group = report_role), colour = col_gridline, linewidth = 1.6) +
  geom_point(aes(colour = series, shape = series), size = 2.6) +
  scale_colour_manual(values = setNames(c(col_accent, col_ink_muted), series_levels)) +
  scale_shape_manual(values = setNames(c(16, 124), series_levels)) +
  labs(x = "ARRS FTE per 10,000 patients", colour = NULL, shape = NULL) +
  chart_theme +
  theme(
    legend.position = "top", legend.justification = "left",
    legend.text = element_text(colour = col_ink_secondary, size = 7.5),
    axis.title.x = element_text(colour = col_ink_secondary, size = 7.5),
    panel.grid.major.y = element_blank()
  )
```

# Limitations

- **A question, not a verdict.** Case-mix, coding practice and local
  context vary in ways this data doesn't capture.
- **ARRS-eligible roles, not confirmed ARRS funding.** We count staff the
  PCN employs in roles the 2026/27 Network Contract DES says ARRS can fund.
  Public data doesn't show how each post is actually paid for. Staff in
  equivalent roles employed directly by practices aren't counted, and some
  PCNs may record ARRS staff under a lead practice or federation instead.
- **Missing or implausible returns are left out, not counted as zero.**
  `{r} n_pcns_excluded` PCNs are removed (no ARRS return, no ARRS roles in
  their return, or a return too implausible to be one PCN's).
- **"Expected" FTE is a national average, not an entitlement.** It's
  fitted from national data on each PCN's age, sex and deprivation mix.
  FTE isn't a true count, so the control limits are an approximation.
- **Our own weights, not the NHS Carr-Hill formula**, whose coefficients
  we couldn't verify. Deprivation weights broadly agree with Mukhtar et
  al. 2018; age weights don't — an open question. Morbidity isn't
  included.
- **No £ figures.** No public source reports ARRS spend below ICB level.
