---
title: "PCN-ARRS diagnostic snapshot"
format:
  typst:
    papersize: a4
    margin:
      x: 1.8cm
      y: 1.6cm
params:
  pcn_code: "U24992" # Heritage PCN — test case; sits on the edge of the 95% band (within range as of the June 2026 build)
---

```{r}
#| label: setup
#| include: false

library(targets)
library(dplyr)
library(ggplot2)
library(tibble)

source_dir <- "../src"
for (f in list.files(source_dir, full.names = TRUE)) source(f)

# Reads pre-built pipeline outputs — run `targets::tar_make()` from the
# project root before rendering this report. Nothing is recomputed here.
store <- "../_targets"
tr <- function(name) tar_read_raw(name, store = store)

national_pcn_list_size <- tr("national_pcn_list_size")
national_pcn_primary_icb <- tr("national_pcn_primary_icb")
national_pcn_workforce <- tr("national_pcn_workforce")
national_staffing_funnel_flags <- tr("national_staffing_funnel_flags")
national_staffing_gap <- tr("national_staffing_gap")
national_pcn_fte_coverage <- tr("national_pcn_fte_coverage")
national_pcn_exclusions <- tr("national_pcn_exclusions")

pcn_code <- params$pcn_code

pcn_name <- filter(national_pcn_list_size, PCN_CODE == pcn_code)$PCN_NAME
icb_name <- filter(national_pcn_primary_icb, PCN_CODE == pcn_code)$ICB_NAME

list_size_row <- filter(national_pcn_list_size, PCN_CODE == pcn_code)
workforce_row <- filter(national_pcn_workforce, PCN_CODE == pcn_code)
funnel_row <- filter(national_staffing_funnel_flags, PCN_CODE == pcn_code)
coverage_row <- filter(national_pcn_fte_coverage, PCN_CODE == pcn_code)

# PCNs with no ARRS return are left out of the funnel entirely (see
# compute_funnel.R::pcn_staffing_data()) — there's no staffing comparison
# to report for them, so refuse to render rather than print an empty page.
if (nrow(funnel_row) == 0) {
  stop(pcn_code, " is not in the national staffing funnel (see ",
       "docs/pcn_exclusions.md) — no report to render.")
}

# Data-coverage figures for the Limitations section.
n_pcns_excluded <- sum(national_pcn_exclusions$scope == "Whole PCN")
n_practices_no_fte <- sum(national_pcn_fte_coverage$n_practices_no_fte)
list_size_no_fte <- sum(national_pcn_fte_coverage$list_size_no_fte)
this_pcn_no_fte <- coverage_row$n_practices_no_fte
fmt_n <- function(x) format(x, big.mark = ",")

this_pcn_coverage_sentence <- if (this_pcn_no_fte > 0) {
  sprintf(
    "For %s, %d of its %d practices (%s patients) had no return and %s left out, so its comparison covers the remaining %s patients.",
    pcn_name, this_pcn_no_fte, coverage_row$n_practices,
    if (this_pcn_no_fte == 1) "is" else "are",
    fmt_n(coverage_row$list_size_no_fte),
    fmt_n(list_size_row$list_size - coverage_row$list_size_no_fte)
  )
} else {
  sprintf("All %d of %s's practices reported staffing.",
          coverage_row$n_practices, pcn_name)
}

# The headline claim a PCN is identified because its staffing is a
# statistical outlier, not because it happens to rank highest — so the
# copy has to read differently depending on whether this PCN actually
# is one (direction points from lower-than-expected FTE for the ARRS
# case; a small minority of PCNs land here — see
# national_staffing_funnel_flags for the national split).
flag_sentence <- dplyr::case_when(
  funnel_row$funnel_flag == "outside_99.8pct" &
    funnel_row$direction == "understaffed_vs_expected" ~
    "That gap is large enough to sit outside the 99.8% range — the strongest, most statistically defensible case for the next ARRS hire.",
  funnel_row$funnel_flag == "outside_95pct" &
    funnel_row$direction == "understaffed_vs_expected" ~
    "That gap sits outside the 95% range where staffing this far below expectation is unlikely to be chance — worth a serious look at the next ARRS hire.",
  funnel_row$funnel_flag == "within_expected_range" ~
    "That's within the range of normal variation nationally — not a statistical outlier on its own.",
  TRUE ~
    "That's actually above what's expected given its patient population — the opposite direction, worth understanding before assuming a staffing gap here."
)

# Palette — dataviz skill's validated reference instance
# (references/palette.md). Swap for a brand palette by changing these
# constants only.
col_accent <- "#2a78d6" # categorical slot 1 (blue)
col_muted <- "#c3c2b7" # baseline/axis gray, used for de-emphasised bars
col_ink_primary <- "#0b0b0b"
col_ink_secondary <- "#52514e"
col_ink_muted <- "#898781"
col_gridline <- "#e1e0d9"
col_good <- "#0ca30c"
col_critical <- "#d03b3b"

chart_theme <- theme_minimal(base_size = 9) +
  theme(
    text = element_text(colour = col_ink_secondary),
    axis.text = element_text(colour = col_ink_muted, size = 7.5),
    axis.title = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = col_gridline, linewidth = 0.3),
    plot.title = element_text(colour = col_ink_primary, size = 10, face = "bold"),
    plot.subtitle = element_text(colour = col_ink_secondary, size = 8),
    legend.position = "none"
  )

pct1 <- function(x) sprintf("%.1f%%", x)
```

# `{r} pcn_name`

`{r} icb_name` &nbsp;·&nbsp; `{r} format(list_size_row$list_size, big.mark = ",")` registered patients &nbsp;·&nbsp; rendered `{r} format(Sys.Date())`

```{r}
#| label: headline
#| echo: false
#| output: asis

cat("::: {.callout-important appearance=\"simple\"}\n\n")
cat(sprintf(
  "**%s is staffed at %.0f%% of what its age profile and deprivation predict.** %.1f FTE in post against an expected %.1f FTE, benchmarked against every PCN in England (Spiegelhalter funnel plot — see below). %s\n\n",
  pcn_name, funnel_row$staffing_ratio_pct, funnel_row$actual_fte, funnel_row$expected_fte,
  flag_sentence
))
cat(":::\n")
```

## At a glance

| Metric | This PCN | What it means |
|---|---|---|
| Staffing vs. expected need | `{r} sprintf("%.0f%%", funnel_row$staffing_ratio_pct)` (`{r} gsub("_", " ", funnel_row$funnel_flag)`) | Actual FTE against what the PCN's age profile and deprivation predict, benchmarked nationally — a PCN is flagged for being a statistical outlier, not for topping a ranking |
| ARRS FTE currently in post | `{r} sprintf("%.1f", workforce_row$arrs_fte)` | Additional-roles staff already funded and working in this PCN |

## Is this staffing gap real, or just noise?

```{r}
#| label: fig-staffing-funnel
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: !expr paste0("Actual vs. expected FTE across all ", nrow(national_staffing_funnel_flags), " PCNs nationally (Spiegelhalter overdispersion-adjusted control limits). ", pcn_name, " marked in blue.")

funnel_data <- national_staffing_funnel_flags |>
  mutate(is_this_pcn = PCN_CODE == pcn_code)

this_row <- filter(funnel_data, is_this_pcn)

# Flip label alignment depending on which half of the (log) x-range the
# highlighted point falls in, so the direct label never runs off the
# panel edge regardless of which PCN this report is parameterised for.
x_mid_log <- mean(log10(range(funnel_data$expected_fte)))
label_hjust <- if (log10(this_row$expected_fte) > x_mid_log) 1.1 else -0.1

ggplot(funnel_data, aes(x = expected_fte, y = staffing_ratio_pct)) +
  geom_hline(yintercept = this_row$od95_lower_pct, colour = col_ink_muted, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od95_upper_pct, colour = col_ink_muted, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od99_8_lower_pct, colour = col_ink_muted, linetype = "dotted", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od99_8_upper_pct, colour = col_ink_muted, linetype = "dotted", linewidth = 0.3) +
  geom_point(aes(colour = is_this_pcn, size = is_this_pcn), alpha = 0.7) +
  geom_text(
    data = this_row, aes(label = pcn_name),
    hjust = label_hjust, size = 2.8, colour = col_ink_primary, fontface = "bold"
  ) +
  scale_colour_manual(values = c(`TRUE` = col_accent, `FALSE` = col_muted)) +
  scale_size_manual(values = c(`TRUE` = 2.6, `FALSE` = 1)) +
  scale_x_log10(
    labels = scales::label_number(scale_cut = scales::cut_short_scale()),
    expand = expansion(mult = c(0.12, 0.12))
  ) +
  chart_theme +
  theme(axis.text = element_text(colour = col_ink_muted, size = 7.5))
```

Dashed lines mark the 95% control band, dotted lines the 99.8% band — both
computed nationally, so a PCN isn't flagged just because it's small and its
staffing ratio is noisy. "Expected" FTE reflects two things about this PCN's
own registered population: its age profile (patients under 15, 15–64, and
65+ each carry a different national-average staffing rate, since older
patients are well established to need more GP time) and its deprivation
profile — a standard indirect-standardisation approach (Spiegelhalter DJ,
"Funnel plots for comparing institutional performance," *Statistics in
Medicine* 2005;24(8):1185-1202). It doesn't need a peer group to make its
point: `{r} pcn_name` is plotted against every PCN in England, on the same
measure, with nothing else mixed in.

This model uses current (2026) national staffing and population data, and
its expected-staffing weights have been checked against peer-reviewed
academic research on GP consultation patterns (Mukhtar et al. 2018, *Br J
Gen Pract*, DOI 10.3399/bjgp18X695981).

## How far is this from similar PCNs?

```{r}
#| label: fig-staffing-gap
#| echo: false
#| fig-width: 6.4
#| fig-height: 1.9
#| fig-cap: !expr paste0("How many more full-time-equivalent (FTE) staff ", pcn_name, " would have if it matched each national reference point. FTE covers all staff groups (GPs, direct patient care and ARRS roles). It is not a headcount, because many posts are part-time. Targets compare this PCN with PCNs that have a similar population; they are not a staffing requirement.")

gap_data <- filter(national_staffing_gap, PCN_CODE == pcn_code) |>
  mutate(
    benchmark = factor(benchmark, levels = rev(levels(benchmark))),
    gap_label = if_else(
      gap_fte > 0,
      sprintf("%.1f FTE  (+%.1f)", target_fte, gap_fte),
      sprintf("%.1f FTE  (already met)", target_fte)
    )
  )
actual_fte_now <- gap_data$actual_fte[1]
x_max <- max(gap_data$target_fte, actual_fte_now)

ggplot(gap_data, aes(y = benchmark)) +
  geom_vline(xintercept = actual_fte_now, colour = col_ink_muted, linewidth = 0.3) +
  geom_segment(
    aes(x = actual_fte, xend = target_fte, yend = benchmark),
    colour = col_accent, linewidth = 2, lineend = "round"
  ) +
  geom_point(aes(x = target_fte), colour = col_accent, size = 2.6) +
  geom_text(aes(x = target_fte, label = gap_label), hjust = -0.18, size = 2.6, colour = col_ink_primary) +
  annotate(
    "text", x = actual_fte_now, y = 4.55, label = sprintf("In post now: %.1f FTE", actual_fte_now),
    hjust = 1.05, size = 2.5, colour = col_ink_secondary
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.02))) +
  coord_cartesian(xlim = c(NA, x_max * 1.22), clip = "off") +
  chart_theme +
  theme(panel.grid.major.y = element_blank())
```

# Limitations

- **A question, not a verdict.** Case-mix, coding practice and local
  context vary in ways this data doesn't capture.
- **Missing or implausible staffing returns are left out, not counted as
  zero.** `{r} n_pcns_excluded` PCNs are removed entirely (no ARRS
  return, or a return too implausible to be one PCN's), as are
  `{r} fmt_n(n_practices_no_fte)` practices (`{r} fmt_n(list_size_no_fte)`
  patients) with no GP or direct-patient-care staff return — their
  patients and staff both. `{r} this_pcn_coverage_sentence`
- **"Expected" FTE is a national average, not an entitlement.** It's
  fitted from national data on each PCN's age, sex and deprivation mix.
  FTE isn't a true count, so the control limits are an approximation.
- **Our own weights, not the NHS Carr-Hill formula**, whose coefficients
  we couldn't verify. Deprivation weights agree closely with Mukhtar et
  al. 2018; age weights don't — an open question. Morbidity isn't
  included.
- **No £ figures.** No public source reports ARRS spend below ICB level.
  Appointment and practice-level detail is available in a follow-up
  conversation.
