---
title: "PRAQTIS diagnostic snapshot"
format:
  typst:
    papersize: a4
    margin:
      x: 1.8cm
      y: 1.6cm
params:
  pcn_code: "U24992" # Heritage PCN — flagged as a staffing outlier (outside 95% band) by the ROS funnel plot
---

```{r}
#| label: setup
#| include: false

library(targets)
library(dplyr)
library(ggplot2)
library(tibble)

source_dir <- "../src"
for (f in list.files(source_dir, full.names = TRUE)) source(f)

# Reads pre-built pipeline outputs — run `targets::tar_make()` from the
# project root before rendering this report. Nothing is recomputed here.
store <- "../_targets"
tr <- function(name) tar_read_raw(name, store = store)

pilot_pcn_list_size <- tr("pilot_pcn_list_size")
pilot_pcn_imd <- tr("pilot_pcn_imd")
pilot_pcn_ros <- tr("pilot_pcn_ros")
pilot_pcn_qof <- tr("pilot_pcn_qof")
pilot_pcn_ppd <- tr("pilot_pcn_ppd")
national_pcn_ros_funnel_flags <- tr("national_pcn_ros_funnel_flags")

pcn_code <- params$pcn_code

pcn_name <- filter(pilot_pcn_list_size, PCN_CODE == pcn_code)$PCN_NAME
icb_name <- "NHS South East London Integrated Care Board"

ros_row <- filter(pilot_pcn_ros, PCN_CODE == pcn_code)
ppd_row <- filter(pilot_pcn_ppd, PCN_CODE == pcn_code)
list_size_row <- filter(pilot_pcn_list_size, PCN_CODE == pcn_code)

# Read from the national table (not a pilot-filtered one) since the
# funnel chart below has to plot the full national spread for its
# control limits to mean anything — same reasoning as PPD's peer group.
ros_funnel_row <- filter(national_pcn_ros_funnel_flags, PCN_CODE == pcn_code)

# The headline claim a PCN is identified because its staffing is a
# statistical outlier, not because it happens to rank highest — so the
# copy has to read differently depending on whether this PCN actually
# is one (direction points from lower-than-expected FTE for the ARRS
# case; a small minority of PCNs land here — see
# national_pcn_ros_funnel_flags for the national split).
ros_flag_sentence <- dplyr::case_when(
  ros_funnel_row$funnel_flag == "outside_99.8pct" &
    ros_funnel_row$direction == "understaffed_vs_expected" ~
    "That gap is large enough to sit outside the 99.8% range — the strongest, most statistically defensible case in the pilot for the next ARRS hire.",
  ros_funnel_row$funnel_flag == "outside_95pct" &
    ros_funnel_row$direction == "understaffed_vs_expected" ~
    "That gap sits outside the 95% range where staffing this far below expectation is unlikely to be chance — worth a serious look at the next ARRS hire.",
  ros_funnel_row$funnel_flag == "within_expected_range" ~
    "That's within the range of normal variation nationally — not a statistical outlier on its own, though still worth reading alongside the peer comparison below.",
  TRUE ~
    "That's actually above what's expected given its patient population — the opposite direction, worth understanding before assuming a staffing gap here."
)

# Palette — dataviz skill's validated reference instance
# (references/palette.md). Swap for a brand palette by changing these
# constants only.
col_accent <- "#2a78d6" # categorical slot 1 (blue)
col_muted <- "#c3c2b7" # baseline/axis gray, used for de-emphasised bars
col_ink_primary <- "#0b0b0b"
col_ink_secondary <- "#52514e"
col_ink_muted <- "#898781"
col_gridline <- "#e1e0d9"
col_good <- "#0ca30c"
col_critical <- "#d03b3b"

chart_theme <- theme_minimal(base_size = 9) +
  theme(
    text = element_text(colour = col_ink_secondary),
    axis.text = element_text(colour = col_ink_muted, size = 7.5),
    axis.title = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = col_gridline, linewidth = 0.3),
    plot.title = element_text(colour = col_ink_primary, size = 10, face = "bold"),
    plot.subtitle = element_text(colour = col_ink_secondary, size = 8),
    legend.position = "none"
  )

pct1 <- function(x) sprintf("%.1f%%", x)
```

# `{r} pcn_name`

`{r} icb_name` &nbsp;·&nbsp; `{r} format(list_size_row$list_size, big.mark = ",")` registered patients &nbsp;·&nbsp; rendered `{r} format(Sys.Date())`

```{r}
#| label: headline
#| echo: false
#| output: asis

cat("::: {.callout-important appearance=\"simple\"}\n\n")
cat(sprintf(
  "**%s is staffed at %.0f%% of what its patient population and deprivation profile predict.** %.1f FTE in post against an expected %.1f FTE, benchmarked against every PCN in England (Spiegelhalter funnel plot — see below). %s\n\n",
  pcn_name, ros_funnel_row$staffing_ratio_pct, ros_funnel_row$actual_fte, ros_funnel_row$expected_fte,
  ros_flag_sentence
))
cat(":::\n")
```

## At a glance

| Metric | This PCN | What it means |
|---|---|---|
| Staffing vs. expected need (ROS funnel) | `{r} sprintf("%.0f%%", ros_funnel_row$staffing_ratio_pct)` (`{r} gsub("_", " ", ros_funnel_row$funnel_flag)`) | Actual FTE against what list size and deprivation predict, benchmarked nationally — a PCN is flagged for being a statistical outlier, not for topping a ranking |
| ARRS FTE currently in post | `{r} sprintf("%.1f", ros_row$arrs_fte)` | Additional-roles staff already funded and working in this PCN |
| Peer Performance Deficit (PPD) | `{r} sprintf("%.1f", ppd_row$ppd)`-point gap | Whether QOF achievement here trails PCNs that face the same list size, deprivation, and disease burden — see below |

## Is this staffing gap real, or just noise?

```{r}
#| label: fig-ros-funnel
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: !expr paste0("Actual vs. expected FTE across all ", nrow(national_pcn_ros_funnel_flags), " PCNs nationally (Spiegelhalter overdispersion-adjusted control limits). ", pcn_name, " marked in blue.")

funnel_data <- national_pcn_ros_funnel_flags |>
  mutate(is_this_pcn = PCN_CODE == pcn_code)

this_row <- filter(funnel_data, is_this_pcn)

# Flip label alignment depending on which half of the (log) x-range the
# highlighted point falls in, so the direct label never runs off the
# panel edge regardless of which PCN this report is parameterised for.
x_mid_log <- mean(log10(range(funnel_data$expected_fte)))
label_hjust <- if (log10(this_row$expected_fte) > x_mid_log) 1.1 else -0.1

ggplot(funnel_data, aes(x = expected_fte, y = staffing_ratio_pct)) +
  geom_hline(yintercept = this_row$od95_lower_pct, colour = col_ink_muted, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od95_upper_pct, colour = col_ink_muted, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od99_8_lower_pct, colour = col_ink_muted, linetype = "dotted", linewidth = 0.3) +
  geom_hline(yintercept = this_row$od99_8_upper_pct, colour = col_ink_muted, linetype = "dotted", linewidth = 0.3) +
  geom_point(aes(colour = is_this_pcn, size = is_this_pcn), alpha = 0.7) +
  geom_text(
    data = this_row, aes(label = pcn_name),
    hjust = label_hjust, size = 2.8, colour = col_ink_primary, fontface = "bold"
  ) +
  scale_colour_manual(values = c(`TRUE` = col_accent, `FALSE` = col_muted)) +
  scale_size_manual(values = c(`TRUE` = 2.6, `FALSE` = 1)) +
  scale_x_log10(
    labels = scales::label_number(scale_cut = scales::cut_short_scale()),
    expand = expansion(mult = c(0.12, 0.12))
  ) +
  chart_theme +
  theme(axis.text = element_text(colour = col_ink_muted, size = 7.5))
```

Dashed lines mark the 95% control band, dotted lines the 99.8% band — both
computed nationally, so a PCN isn't flagged just because it's small and its
staffing ratio is noisy. "Expected" FTE is the national average FTE per unit
of weighted need (list size × IMD score — ROS's own numerator, see
`src/compute_ros.R`) applied to this PCN's own list. It doesn't need a peer
group to make its point: `{r} pcn_name` is plotted against every PCN in
England, on the same measure, with nothing else mixed in.

## Is the gap real, or just this PCN?

`{r} pcn_name` is also behind PCNs that face the closest thing to the same
challenges nationally (matched on list size, deprivation, and disease-register
prevalence): those PCNs hit a median
**`{r} pct1(ppd_row$peer_median_qof_achievement_pct)`** on QOF achievement
against `{r} pcn_name`'s **`{r} pct1(ppd_row$own_qof_achievement_pct)`** — a
**`{r} sprintf("%.1f", ppd_row$ppd)`-point gap** against `{r} ppd_row$n_peers`
peer PCNs. That's a second, independent signal pointing the same direction as
ROS: this isn't one metric's fluke, there's real headroom here. This finding
is a question worth asking, not a verdict — see *Limitations* for what a
peer-group comparison can and can't prove.

# Limitations

- **This is a question, not a verdict.** Every gap above is worth a
  conversation, not a conclusion — case-mix, coding practice, and local
  context all vary in ways this data doesn't capture.
- **ARRS FTE gaps are treated as zero, not missing.** A handful of PCNs have
  no rows in the PCN Workforce file; ROS treats that as zero ARRS staff — the
  conservative read, since understating ARRS FTE only raises ROS — rather
  than omitting the PCN, but the true figure could be higher than shown (see
  `src/compute_ros.R`).
- **"Expected" FTE is a national average, not a funding entitlement.** It's
  built by applying the national average staffing rate (FTE per unit of
  list size × IMD) to this PCN's own list — the same indirect-standardisation
  logic as NHS SMR-style funnel plots, but FTE is fractional, not a genuine
  count of discrete events, so the Poisson-based control limits are an
  approximation, not an exact statistical fit.
- **PPD's peer group is our own construction**, replicating NHS England's
  published CCG-similarity method — not an official government comparator.
- **No per-PCN £ figure for ARRS spend or underspend exists publicly**
  below ICB level — nothing here should be read as a costed claim.
- **QOF disease-register prevalence** (used to build the peer group) is a
  relative similarity index across 21 conditions, not a clinical prevalence
  rate — it double-counts multimorbid patients by design.
- **DNA (Did Not Attend) rates are not yet in this pipeline** — CLAUDE.md's
  Phase 3 spec calls for a DNA benchmark and it isn't here yet.
- **This report is a first draft**, rendered against one pilot PCN as a
  test case. Page count, layout, and section order are all still open.
